---
title: "Fitting Regression Models"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Chapter 10 --- Fitting Regression Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source('_format.R')
```


Chapter 10 of the text discusses fitting regression models.  Thankfully, R's support for such models is very good.  Again, we must be carefult to set the contrasts of unordered factors to "contr.sum" as follows:
```{r}
options(contrasts=c(unordered='contr.sum', ordered='contr.poly'))
```


# Example 10.1


The example investigates the relationship of reaction temperature and catalyst feed rate to the viscosity of a polymer. First, let's do the regression through matrix operations.  Define the $X$ matrix using the `model.matrix` command:

```{r}
X <- model.matrix(~ReactionTemperature + CatalystFeedRate, data=Table10.2)
print(X)
```

\noindent The response $y$ is just the `Viscosity` column:

```{r}
y <- Table10.2$Viscosity
print(y)
```

\noindent Now we solve $(X^\mathrm{T} X)^{-1} X^\mathrm{T} y$.  We can find the inverse $(X^\mathrm{T} X)^{-1}$ with `solve(t(X) %*% X)` and then multiply by $X^\mathrm{T} y$, but it's numerically more stable to supply $X^\mathrm{T} y$ as the second argument to `solve`:
```{r}
beta.hat <- solve(t(X) %*% X, t(X) %*% y)
print(beta.hat)
```

\noindent Now, to do this using `lm` is far simpler:
```{r}
model <- lm(Viscosity ~ ReactionTemperature + CatalystFeedRate, data=Table10.2)
print(summary(model))
```

Table 10.3 gives many types of residuals and diagnostic measures, many of which we'll create below:

```{r}
Table10.3 <- data.frame(
	'y'=Table10.2$Viscosity,
	'Predicted'=predict(model),
	'Residual'=resid(model),
	'hii'=hatvalues(model),
	'StudentizedResids'=rstudent(model),
	'CooksD'=cooks.distance(model)
)

```
```{r echo=FALSE}
kable(Table10.3) %>% kable_styling()
```

-------------


# Example 10.6

Consider the viscosity data from Table 10.2.  If we wish to test $H_0: \beta_2 = 0$ versus $H_1: \beta_2 \ne 0$ by partial $F$-test we can use the `anova` function:

```{r}
model01 <- lm(Viscosity ~ ReactionTemperature, data=Table10.2)
model012 <- lm(Viscosity ~ ReactionTemperature + CatalystFeedRate, data=Table10.2)
print(anova(model01, model012))
```

-------------

# Example 10.7

Parameter confidence intervals can be retrieved with the `confint` function:
```{r}
model <- lm(Viscosity ~ ReactionTemperature + CatalystFeedRate, data=Table10.2)
confint(model)
```
\noindent We can also get other intervals.  For example, the prediction interval for the response over `ReactionTemperature` at `CatalystFeedRate` set to 10.5 is given below:  
```{r fig.align='center', fig.width=5, fig.height=5}
x1 <- seq(from=80, to=100, length.out=100)
x2 <- rep(10.5, 100)
y <-predict(model, newdata=data.frame('ReactionTemperature'=x1, 'CatalystFeedRate'=x2), interval="confidence") 
plot(x1, y[,1], type='l', ylim=c(2248, 2434), main='Prediction Interval', xlab='Reaction Temperature', ylab='Viscosity')
lines(x1, y[,2], lty=2)
lines(x1, y[,3], lty=2)
```
