---
title: "12. Robust Parameter Design and Process Robustness Studies"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{12. Robust Parameter Design and Process Robustness Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source('_format.R')
```

Chapter 12 in the text introduces robust parameter designs.  These designs allow the experimenter to investigate situations where not all of the factors can be easily controlled.    


# Example 12.1


Recall the filtration experiment from Example 6.2.  Assume that temperature (`z`) is difficult to control, while pressure (`x1`), concentration (`x2`), and stirring rate (`x3`) are easy to control. We fit the following:
```{r}
df <- Table6.10
colnames(df) <- c('z', 'x1', 'x2', 'x3', 'y', 'Block')
model <- lm(y ~ z * (x1 + x2 + x3), data=df)
print(summary(model))
```
\noindent Since $x_1$ is not significant let's refit:
```{r}
model <- lm(y ~ z * (x2 + x3), data=df)
print(summary(model))
```
\noindent We can pick those terms that do not involve $z$ to find $\mathbb{E}_z[y(\mathbf{x}, z)]$:
$$
\mathbb{E}_z[y(\mathbf{x}, z)] = 70.06 + 4.94 x_2 + 7.31 x_3~.
$$
Next, $\mathrm{Var}_z\bigl[y(\mathbf{x}, z)\bigr]$ is  $$
\mathrm{Var}_z\bigl[y(\mathbf{x}, z)\bigr]
=
[10.812 - 9.063 x_2 + 8.312 x_3]^2 \sigma_z^2 + \sigma^2 ~.
$$

Holding temperature $z=0$ we can plot the contours (blue) of the expected value, then plot the contours (orange) of the variance on top of that:
```{r fig.align='center', fig.width=5, fig.height=5}
x2 <- seq(from=-1, to=1, length.out=100)
x3 <- seq(from=-1, to=1, length.out=100)
z <- matrix(predict(model, cbind(
	'z'=rep(0, 100^2),
	expand.grid('x2'=x2, 'x3'=x3)
)), ncol=100)

poe.coefs <- (
	model.matrix(
		~ z * (x2 + x3),
		data=cbind(
			'z'=rep(1, 100^2),
			expand.grid('x2'=x2, 'x3'=x3)
		)
	) 
	%*% 
	matrix(c(0, coef(model)['z'], 0, 0, coef(model)[c('z:x2','z:x3')]))
)^2
z.poe <- matrix(sqrt(poe.coefs), nrow=100)

contour(x2, x3, z, col='blue', xlab='Concentration', ylab='StirringRate')
contour(x2, x3, z.poe, col='orange', add=TRUE)
```

-------------

# Example 12.2

Consider the semiconductor manufacturing experiment data in Table 12.3. We fit the following model:
```{r}
df <- Table12.3[,-1]
model <- lm(y ~ (z1 + z2 + z3) * (x1 + x2) + x1:x2 + I(x1^2) + I(x2^2), data=df)
print(summary(model))
```

\noindent We can pick those terms that do not involve $z_i$ to find $\mathbb{E}_z[y(\mathbf{x}, z)]$:
$$
\mathbb{E}_z[y(\mathbf{x}, z)] = 30.365 -2.921 x_1 - 4.129 x_2 + 2.869 x_1 x_2 + 2.596 x_1^2 + 2.183 x_2^2~.
$$
Next, $\mathrm{Var}_z\bigl[y(\mathbf{x}, \mathbf{z})\bigr]$ is  \begin{align}
\mathrm{Var}_{\mathbf{z}}\bigl[y(\mathbf{x}, \mathbf{z})\bigr]
&=
[2.731 - 0.269 x_1  + 2.006 x_2]^2 \sigma_{z_1}^2 \\
&\phantom{=}~+
[-2.331 + 0.894 x_1 -1.431 x_2]^2 \sigma_{z_2}^2 \\
&\phantom{=}~+
[2.331 + 2.581 x_1 + 1.556 x_2]^2 \sigma_{z_3}^2 \\
&\phantom{=}~+
\sigma^2 ~.
\end{align}

Letting $\sigma_{z_i}^2 = 1$ we plot the contours of mean (blue) and variance (orange) as follows:
```{r fig.align='center', fig.width=5, fig.height=5}
x1 <- seq(from=-1, to=1, length.out=100)
x2 <- seq(from=-1, to=1, length.out=100)

mean.df <- cbind(
	data.frame(
		'z1'=rep(0, 100^2),
		'z2'=rep(0, 100^2),
		'z3'=rep(0, 100^2)
	),
	expand.grid('x1'=x1, 'x2'=x2)
)

z <- matrix(predict(model, mean.df), ncol=100)

var.df <- cbind(
	data.frame(
		'z1'=rep(1, 100^2),
		'z2'=rep(1, 100^2),
		'z3'=rep(1, 100^2)
	),
	expand.grid('x1'=x1, 'x2'=x2)
)

X.m <- model.matrix( ~ (z1 + z2 + z3) * (x1 + x2) + x1:x2 + I(x1^2) + I(x2^2), data=var.df)
b <- coef(model)
z1.terms <- matrix(c(0, b[2], 0, 0, 0, 0, 0, 0, b[9], b[10], 0, 0, 0, 0, 0))
z2.terms <- matrix(c(0, 0, b[3], 0, 0, 0, 0, 0, 0, 0, b[11], b[12], 0, 0, 0))
z3.terms <- matrix(c(0, 0, 0, b[4], 0, 0, 0, 0, 0, 0, 0, 0, b[13], b[14], 0))

poe.coefs <- sqrt((X.m %*% z1.terms)^2 + (X.m %*% z2.terms)^2 + (X.m %*% z3.terms)^2)
z.poe <- matrix(poe.coefs, nrow=100)

contour(x1, x2, z, col='blue', xlab='x1', ylab='x2')
contour(x1, x2, z.poe, col='orange', add=TRUE)
```
